---
title: "DSC 520 Week 5 Correlation Assignment"
author: "Dan Wiltse"
date: "Jan 5 2020"
output:
  pdf_document: default
  word_document: default
---
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# R Markdown
# a. Use R to calculate the covariance of the Survey variables and provide an explanation #of why you would use this calculation and what the results indicate.

---
params: 
  filename = "student-survey.csv"

---
```{r}
cov(student.survey)

```


#b. Examine the Survey data variables. What measurement is being used for the #variables? Explain what effect changing the measurement being used for the variables #would have on the covariance calculation. Would this be a problem? Explain and #provide a better alternative if needed.
#c. Choose the type of correlation test to perform, explain why you chose this test, and #make a prediction if the test yields a positive or negative correlation?
#d. Perform a correlation analysis of: 
# 1)All variables
#2) A single correlation between two of the variables
#3) Repeat your correlation test in step 2 but set the confidence interval at 99%
#4) Describe what the calculations in the correlation matrix suggest about the relationship between the variables. Be specific with your explanation.

#e. Calculate the correlation coefficient and the coefficient of determination, describe what you conclude about the results.
#f. Based on your analysis can you say that watching more TV caused students to read less? Explain.
#g. Use TV Time and Happiness while controlling for Gender and perform a partial correlation. Explain how this changes your interpretation and explanation of the results.


## Including Plots

You can also embed plots, for example:
  
  ```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## References

